### 2014-11-05 abstract.tex
More corrections.

Declarative programming is good for parallel programming but makes it hard for
the programmer to fine tune scheduling and data layout of the program.
some programs require more information in order to execute better.
with our approach we permit the programmer to write declarative programs
but then optimize them later using sensing facts and action facts that
understand and change how the programs are executed.

Linear Meld is a concurrent forward-chaining linear logic programming
language where logical facts can be asserted and retracted in a
structured way. In Linear Meld, a program is seen as a database of
logical facts and a set of derivation rules. The database of facts is
partitioned by the nodes of a graph structure which leads to

>>>>

Declarative programming has been hailed as a promising approach to parallel
programming since it hides the implementation details of parallelism away from the
programmer. However, some control is lost when code is
automatically parallelized, leaving the programmer with very little
opportunities for fine tuning programs.
We extended Linear Meld, a concurrent forward-chaining linear logic programming
language, with coordination facts that allow the programmer to change
computation scheduling and data layout. Our approach allows the programmer to
write declarative parallel programs and then optionally use coordination facts
to fine tune programs, while keeping the program fully declarative and amenable
to correctness proofs. We show several application examples, experimental
results and some advantages of our novel approach.

### 2014-11-05 coordination.tex
Add more text in MiniMax.
More corrections.

We introduce the concept of \emph{coordination facts} that allow the programmer
to change how the run time schedules nodes and partitions the nodes among
threads of execution. Coordination facts can be used in the body or head of the
rules to allow the programmer to reason and change how execution is to be done.
In this fashion, we keep the language declaratively because we reason logically
about the state of execution, without the need to introduce extra-logical
operators into the language that would reduce problems when attempting to prove
properties about programs.

There are two kinds of coordination facts. The first kind of coordination facts are
called \emph{sensing facts} and are used to sense information about data
placement (where is the node being executed) or scheduled (what is the priority
of the node). The second kind of coordination facts are \emph{action facts} that,
when derived, are consumed to apply a coordination operation during execution.
These logical rules that use coordination are implemented efficiently in the run
time system and are hidden from the programmer.

\subsection{Scheduling Facts}

We can use action facts to change the order in which nodes are evaluated by adding
\emph{priorities}. Node priority works at the thread level
so that each thread can make a local decision about which node of the graph to run next.
Note that, by default, nodes are picked arbitrarily (a FIFO approach is used).

We have two kinds of priorities: a \emph{temporary priority} and a \emph{default
priority}. A default priority is used indefinitely, while a temporary priority
is only used once, so that when node with a temporary priority $P$ is picked to
run, the priority will default back to $D$, the default priority. Initially,
all nodes have a default priority of $0$.

@@

\begin{itemize}
   \item \texttt{type linear set-priority(node, float)}: This sets the
   temporary priority of a node. If the node already has a priority, we only
   change the priority if the new one is higher priority. The programmer can
   decide if priorities are to be ordered in ascending or descending order.
   \item \texttt{type linear action add-priority(node, float)}: Increments,
   temporarily, the current priority of the node.
   \item\texttt{type linear schedule-next(node)}: The work will fetch the
   highest priority node's priority $P$ from its set of nodes and set the
   action's argument node's priority as $P + 1.0$. If using the priorities in
   ascending order, we pick the lowest priority and subtract $1.0$.
   \item\texttt{type linear set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \texttt{type linear action stop-program(node)}: Immediately stops the
   execution of the whole program.
\end{itemize}

LM provides the sensing fact \texttt{type linear priority(node, node, float)},
where the second argument is the target node and the third argument is the
node priority. Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are consumed, they are re-derived automatically,
except if the programmer explicitly tells the compiler otherwise. 
Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
value of \texttt{priority} facts, but this done seamlessly by the runtime
system.

@@

\begin{itemize}
   \item \texttt{type linear cpu-id(node, node, int)}: The third argument
   indicates the thread where the node of the second argument is currently running.
   \item \texttt{type linear moving(node)}: Fact available if the node in the
   second argument is allowed to move between threads.
   \item \texttt{type linear static(node, node)}: Fact available if the node in
   the second argument is not allowed to move between threads.
\end{itemize}

>>>>

We introduce the concept of \emph{coordination facts} that allow the programmer
to change how the run time schedules nodes and partitions the nodes among
threads of execution. Coordination facts can be used in the body or head of the
rules to allow the programmer to reason and change how execution is to be done.
In this fashion, we keep the language declarative because we reason logically
about the state of execution, without the need to introduce extra-logical
operators into the language that would introduce problems when attempting to prove
properties about programs.

There are two classes of coordination facts. The first class of coordination facts are
called \emph{sensing facts} and are used to sense information about the
underlying runtime system, including node placement and node scheduling.
The second kind of coordination facts are \emph{action facts} that are deleted
to apply a coordination operation in the runtime system.

\subsection{Scheduling Facts}\label{sec:fifo}

We can use action facts to change the order in which nodes are evaluated by adding
\emph{priorities}. Node priority works at the thread level
so that each thread can make a local decision about which node to execute next.
Note that, by default, nodes are picked using a FIFO approach, because that
tends to work better for most programs since older nodes tend to have more facts
to be used.

We have two kinds of priorities: a \emph{temporary priority} and a \emph{default
priority}. A temporary priority momentarily changes the default priority $D$ of a
node, so that once the node is done, the priority will default back to $D$.
Initially, all nodes have a default priority of $0$.

@@

\begin{itemize}
   \item \texttt{type linear set-priority(node, float)}: This sets the
   temporary priority of a node. If the node already has a priority, we only
   change the priority if the new one is higher. The programmer can
   decide if priorities are to be ordered in ascending or descending order.
   \item \texttt{type linear action add-priority(node, float)}: Increases,
   temporarily, the priority of the node.
   \item\texttt{type linear schedule-next(node)}: The work will fetch the
   highest priority node's priority $P$ from its set of nodes and set the
   action's argument node's priority as $P + 1.0$. If using the priorities in
   ascending order, we pick the lowest priority and subtract $1.0$.
   \item\texttt{type linear set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \texttt{type linear action stop-program(node)}: Immediately stops the
   execution of the whole program.
\end{itemize}

LM provides the sensing fact \texttt{priority(A, B, P)} in order to sense the
priority \texttt{P} of node \texttt{B} from node \texttt{A}.
Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are deleted, they are re-derived automatically.
Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
value of \texttt{priority} facts, but this done automatically by the runtime
system.

@@

\begin{itemize}
   \item \texttt{type linear cpu-id(node, node, int)}: The third argument
   indicates the thread where the node of the second argument is currently running.
   \item \texttt{type linear moving(node, node)}: Fact available if the node in the
   second argument is allowed to move between threads.
   \item \texttt{type linear static(node, node)}: Fact available if the node in
   the second argument is not allowed to move between threads.
\end{itemize}

### 2014-11-05 experiments.tex
More corrections.

\subsection{Absolute runtime}

@@

Fig.~\ref{results:comparison1}

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{\includegraphics[width=9cm]{results/comparison1.png}}
   \end{center}
   \caption{Measuring the overhead of coordination mechanisms. The \textbf{Old}
   bars represent the performance of the original virtual machine, while
   \textbf{New} is the relative performance of the new version that includes
   coordination mechanisms. Most programs show little to no degradation in
   performance.}
   \label{results:comparison1}
\end{figure}

>>>>

@@

Fig.~\ref{results:comparison1} shows the comparison between the original virtual
machine and the new virtual machine with coordination mechanisms. All the
programs benchmarked do not use any kind of coordination. As expected, there
is no degration of performance by adding coordination mechanisms.

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{\includegraphics[width=9cm]{results/comparison1.png}}
   \end{center}
   \caption{Measuring the overhead of coordination mechanisms. The \textbf{Old}
   bars represent the performance of the original virtual machine, while
   \textbf{New} is the relative performance of the new version that includes
   coordination mechanisms.}
   \label{results:comparison1}
\end{figure}

### 2014-11-05 heat-transfer.tex
More corrections.

Randomized and approximation
algorithms can obtain significant benefits from coordination directives because although the final
program results will not be exact, they follow important statistical properties and can be computed faster.
Examples of such programs are Loopy Belief
Propagation~\cite{Gonzalez+al:aistats09paraml} (shown in the next section) and
Heat Transfer~(HT).

In HT, we have a graph of nodes that exchange heat values between them. The goal
of the program is to exchange heat to a point where new heat value $\delta = |H_i -
H_{i-1}| \le \epsilon$ for all nodes of the graph. The algorithm works
asynchronously since heat values can be updated by using new partial information
coming from neighbor nodes. This increases parallelism since nodes do not need
to synchronize if we had to compute heat values using iterations.

>>>>

In the Heat Transfer algorithm~(HT), we have a graph where heat values are
exchanged between nodes. The goal of the program is to exchange heat to a point
where new heat value $\delta = |H_i -
H_{i-1}| \le \epsilon$ for all nodes of the graph. The algorithm works
asynchronously since heat values can be updated by using new partial information
coming from neighbor nodes. This increases parallelism since nodes do not need
to synchronize if we had to compute heat values using iterations.

### 2014-11-05 implementation.tex
More corrections.

The implementation of LM includes a compiler and a virtual machine that runs
byte-code. The virtual machine uses 32 registers for
operations and executes procedures to iterate over the database in order to match facts.
The implementation of the original machine is described in ~\cite{cruz-ppdp14}.
Here, we resume the most important details of that paper, while focusing on the
changes for efficiently implementing coordination.

@@

The compiler translates each rule to a procedure and a list of facts that need
to exist to satisfy the rule. This procedure is executed by the virtual
machine whenever we have enough facts to satisfy the rule. The procedure
loops over all possible combinations of the rule, retrieving facts from the
database, performing join operations and then consuming and deriving facts.
Optimizations such as join optimizations (to allow rule filtering invalid
combinations) and fact updates are implemented by the compiler. The latter
transforms fact derivations of the same predicate to a simple update operation,
avoiding a deallocation operation followed by an allocation of the new fact.

\subsection{Node partitioning}

The compiler builds a representation of the graph by inspecting axioms of the
program. Next, it orders the nodes of the graph using a breadth-first
topological sort. This node ordering is written to the byte-code file and then
used to initially partition nodes across available threads.

@@

Coordination directives are compiled in two different ways, depending whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of returning something. While these are still facts from the point of
view of the language, they are consumed immediatelly in order to apply the
operation, instead of being derived like any other regular fact. This is more
efficient since we do not need to create a new fact.

@@

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible
for executing a subset of nodes. Nodes are either inactive, with no new facts,
or active, where they will be placed in the corresponding thread's queue.
Threads do useful work by processing the nodes in their queues. Whenever a
thread becomes idle, it attempts to steal nodes from a random thread into its
own queues. If unsuccessful, the thread synchronizes with the other threads to
become inactive.

@@

Each node is protected by a main spinlock that allows threads to change node
attributes: incoming facts, owner thread, node state and locality information.
There is also an internal spinlock that is locked whenever the node enters into
the \textbf{working} state. If a thread sends facts to a node placed in
another thread, it attempts to lock the internal lock in order to update the
indexing structures of the node, otherwise it adds the facts to the list of
incoming facts that are later processed by the owner thread.

>>>>

The implementation of LM includes a compiler and a virtual machine that runs
byte-code. The virtual machine uses 32 registers for
operations and executes procedures to iterate over the database in order to match facts.
The implementation of the original machine is described in ~\cite{cruz-ppdp14}.
Here, we review the most important details of that paper, while focusing on the
changes for efficiently implementing coordination.

@@

The compiler translates each rule to a procedure and a list of facts that need
to exist to satisfy the rule. This procedure is executed by the virtual
machine whenever we have enough facts to satisfy the rule. The procedure
loops over all possible combinations of the rule, retrieving facts from the
database, performing join operations and then consuming and deriving facts.
Optimizations such as join optimizations (to support efficient rule filtering)
and fact updates are implemented by the compiler. The latter
transforms fact derivations of the same predicate to a simple update operation,
avoiding a deallocation operation followed by an allocation of the new fact.

@@

Coordination directives are compiled in two different ways, depending on whether they
appear in the body or in the head of the rule. Coordination facts in the body
are compiled into special instructions that inspect the state of the virtual
machine. For example, the directive \texttt{node-priority} will inspect the
target node, retrieve the current priority and assign the priority to a
register. Coordination facts in the head of the rule are also implemented as
special instructions of the virtual machine, but they perform some action,
instead of returning something. While these are still facts from the point of
view of the language, they are deleted immediatelly in order to apply the
operation, instead of being derived like any other regular fact. This is more
efficient since we do not need to create a new fact.

@@

The virtual machine is implemented in C++11 and uses the threading system from
the standard library to implement multithreading. Each thread is responsible
for executing a subset of nodes. Nodes are either inactive, with no new facts,
or active, where they will be placed in the corresponding thread's queue.
Threads do useful work by processing the nodes in their queues. Whenever a
thread becomes idle, it attempts to steal nodes from a random thread into its
own queues. If sucessful, the thread becomes idle and waits for program
termination while periodically attempting to steal nodes.

@@

Each node is protected by a main spinlock that allows threads to change node
attributes: incoming facts, owner thread, node state and locality information.
There is also an internal spinlock that is locked whenever the node enters into
the \textbf{working} state. If a node sends facts to node is placed in
another thread, the current thread first attempts to lock the internal lock in order to update the
indexing structures of the target node, otherwise it adds the facts to the list of
incoming facts that are later processed by the owner thread.

### 2014-11-05 intro.tex
More corrections.

Writing parallel programs in sequential languages is hard because manipulating
shared state using multiple threads may result in data race
conditions. Such issues are handled with low level constructs such as locks and
condition variables, requiring a fair amount of effort to get right.
Declarative programming has been hailed as a solution to this issue, since the
problem of implementing the details of parallelism is moved away from the programmer to the
compiler and runtime of the language. The programmer writes code
without having to deal with parallel programming constructs and the compiler
automatically parallelizes the program in order to take advantage of
multiple threads of execution.
The programming paradigm has been adopted with huge success in domain specific
languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
Although more general declarative languages have yet to be as successful, the
future looks promising for this particular approach.

@@

In this paper, we extend Linear Meld (LM), a declarative language by
\cite{cruz-iclp14,cruz-ppdp14}, with coordination facts that give programmer control over
scheduling and data placement. LM is a linear logic programming language
intended for programs over graph structures and with
support for structured manipulation of mutable state using 
linear logic~\cite{girard-87}. In LM, the nodes of the graph perform computation
through derivation of rules and communicate with other nodes when logical facts
are derived.

The operational semantics of LM allow for plenty of
non-determinism since nodes execute independently and can be executed with
any specific order. Since LM uses mutable state, the order in which nodes are
executed may be crucial in terms of performance. We extended the language with
coordination facts that can be used as a regular fact in the program logic to
improve program execution.
The first kind of coordination facts are
called \emph{sensing facts} and are used to sense information about data
placement (where is the node being executed). This allows the programmer to
write logical rules that depend on the current state of the program. The second
kind of coordination facts are \emph{action facts} that, when deriving, will be
consumed to apply a scheduling operation during execution. This allows the
programmer to prioritize node computation or place nodes in different threads.
Both kinds of facts are \emph{linear facts} since they are consumed and/or change.

To the best of our knowledge, this is the first time that a declarative language allows 
control over execution without a significant change in the language itself and
without using non-declarative operations. This proves specially important since
programs can proven to be correct even in the presence of coordination facts.
Our other two main contributions are as follows: a novel way to coordinate
declarative programs through user control that introduce measurable performance
improvements; several application where we add programmer control to declarative
programs in order to make them perform better.

>>>>

Writing parallel programs in sequential languages is hard because manipulating
shared state using multiple threads may result in data race
conditions. Such issues are handled with low level constructs such as locks and
condition variables, requiring a fair amount of effort to get right.
Declarative programming has been hailed as a solution to this issue, since the
problem of implementing the details of parallelism is moved away from the programmer to the
compiler and runtime of the language. The programmer writes code
without having to deal with parallel programming constructs and the compiler
automatically parallelizes the program in order to take advantage of
multiple threads of execution.
The programming paradigm has been adopted with huge success in domain specific
languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
Although general declarative languages have yet to be as successful, the
future looks promising for this particular approach.

@@

In this paper, we extend Linear Meld (LM), a declarative language by
\cite{cruz-iclp14,cruz-ppdp14}, with coordination facts that give programmer
control over scheduling and data placement. LM is a linear logic programming
language designed for programs that operate on graphs.  The use of linear
logic~\cite{girard-87} allows support for structured manipulation of mutable
state. In LM, computation is divided so that each node of the graph computes
independently and is allowed to communicate with other nodes.
Both computation and communication happen through derivation of logical rules.

Given that LM uses mutable state, the order in which nodes are executed may be
crucial in terms of performance. We extended the language with coordination
facts that can be used as a regular fact in the program logic to improve program
execution.  The first kind of coordination facts are called \emph{sensing facts}
and are used to sense information about the system the program is running on.
This allows the programmer to write logical rules that depend on the current
state of the program and the system. The second kind of coordination facts are
\emph{action facts} that when deleted in rules are used to apply a scheduling
operation during execution. This allows the programmer to prioritize node
computation or place nodes in different threads.

To the best of our knowledge, this is the first time that a declarative language allows 
control over execution without a significant change in the language itself and
without using non-declarative operations. This is specially important since
programs can proven to be correct even in the presence of coordination facts.
In order to prove our thesis, we first give a brief overview of the
base language, including a small example. Next, in
Section~\ref{sec:coordination} we introduce the available coordination facts,
followed by Section~\ref{sec:implementation}, where we describe the
implementation of the runtime required to implement the desired coordination
mechanisms. Afterwards, in Section~\ref{sec:applications} we present several applications where we add
programmer control to declarative programs in order to make them perform better.

### 2014-11-05 language.tex
More corrections.

Linear Meld (LM) is a forward-chaining linear logic programming
language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
Programs are written as a set of rules of the form
\texttt{a(X), b(Y) -o c(Z)} and are read as following: if \texttt{a(X), b(Y)} (the body of
the rule) is satisfied using some facts \texttt{a(X)} and \texttt{b(Y)} then
\texttt{c(Z)} (the head of the rule) is true.
The facts used during rule derivation are placed in a \emph{database of facts}.
Logical facts are composed of a predicate and a tuple of concrete values,
representing the arguments.
Since LM uses linear logic as its foundation, rule derivations
may delete facts used in the body of the rule. Program execution starts by
adding the \emph{axioms} (the initial facts) of the program to the database.
Next, the rules are recursively applied and the database is updated with new and
deleted facts. When nore more rules are applicable, the program terminates.

LM has been designed for writing graph-based programs. LM accomplishes this by
partitioning the logical facts by the first argument (the \emph{node}) and by
restricting the body of every rule to only refer to the same first argument.
This allows rules to be executed using only facts of a single node. Although the
body of rules is restricted, the head of the rule may refer to any other node as
long as the node variable is refered somewhere in the body of the rule. This allows
derivations of facts in other nodes. The whole database is then composed of smaller
databases (the nodes) and rule derivations allows smaller databases to derive
facts in other databases, creating edges in the graph.

@@

The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
with the declaration of the predicates. Predicates specify the kinds of facts
used in the program. The first predicate, \texttt{edge}, specifies edges between
nodes of the graph, where the third argument represents the weight of the edge.
The \texttt{route} modifier informs the compiler about the structure of the
program graph. The predicates \texttt{shortest} and \texttt{relax} are specified
as linear facts and thus are deleted when used in rules.
The idea of the algorithm is to compute the shortest distance from node
\texttt{@1} to all other nodes in the graph. Every node has a \texttt{shortest}
fact that is improved with new \texttt{relax} facts.
Lines 5-9 declare the axioms of the program: \texttt{edge} facts describe the
graph; \texttt{shortest(A, +00, [])} is the initial shortest distance (infinity)
for all nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm with the
initial distance from \texttt{@1} to \texttt{@1}.

@@

\begin{figure*}[ht]
\begin{center}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}
\end{center}
\caption{Graphical representation of the SSSP program. Figure (a) represents the
   program after propagating initial distance at node \texttt{@1}, followed by
   Figure (b) where the first rule is applied in node \texttt{@3}. Figure (c)
   represents the state of the final program, where all the shortest paths
   have been computed.}
\label{fig:shortest_path_program}
\end{figure*}

>>>>

Linear Meld (LM) is a forward-chaining linear logic programming
language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
Programs consist of a set of \emph{rules} and a \emph{database of facts}.
Rules such as \texttt{a(X), b(Y) -o c(X, Y)} can be read as follows: if fact \texttt{a(X)}
and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
the database. The expression \texttt{a(X), b(Y)} is called the \emph{body} of
the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
A fact is a predicate and its associated tuple of values, representing the
arguments. Since LM uses linear logic as its foundation, we distinguish between
\emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
process of deriving rule, while persistent facts are not.
Program execution starts by adding the \emph{axioms} (the initial facts) of the program to the database.
Next, the rules are recursively applied and the database is updated with new and
deleted facts. When no more rules are applicable, the program terminates.

LM has been designed for writing programs that operate on graphs. LM partitions
the database by using the first argument of each fact. The first argument is
typed as a \emph{node} and represents a node in the graph. For instance, the
fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while fact \texttt{p(@2)}
is stored in node \texttt{@2}. LM restricts
the body of every rule to only refer to the same node so that nodes can derive
rules independently. Although the body is restricted, the head of the
rule may refer to any node as long as the node is refered somewhere in the
body. This allows communication between nodes during rule derivation, since a
node must send a fact to another node.
Rule restrictions in turn make LM implicitly parallel because nodes are able to
compute independently as long as they have new facts to be applied in rules.
This makes LM non-deterministic since nodes can be picked to run in any order,
affecting which rules are applied and which facts are deleted or derived.

@@

The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
with the declaration of the predicates. Predicates specify the kinds of facts
used in the program. The first predicate, \texttt{edge}, is a persistent
predicate that describes the
relationship between the nodes of the graph, where the third argument represents
the weight of the edge
(the \texttt{route} modifier informs the compiler about the structure of the
program graph). The predicates \texttt{shortest} and \texttt{relax} are specified
as linear facts and thus are deleted when deriving new facts.
The idea of the algorithm is to compute the shortest distance from node
\texttt{@1} to all other nodes in the graph. Every node has a \texttt{shortest}
fact that is improved with new \texttt{relax} facts.
Lines 5-9 declare the axioms of the program: \texttt{edge} facts describe the
graph; \texttt{shortest(A, +00, [])} is the initial shortest distance (infinity)
for all nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm with the
initial distance from \texttt{@1} to \texttt{@1}.

@@

\begin{figure*}[ht]
\begin{center}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
  \hspace{0.4cm}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
  \hspace{0.4cm}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}
\end{center}
\caption{Graphical representation of the SSSP program. Figure (a) represents the
   program after propagating initial distance at node \texttt{@1}, followed by
   Figure (b) where the first rule is applied in node \texttt{@3}. Figure (c)
   represents the state of the final program, where all the shortest paths
   have been computed.}
\label{fig:shortest_path_program}
\end{figure*}

### 2014-11-05 minimax.tex
More corrections.

The algorithm proceeds by building a game tree, where each tree node represents a
game state and the children represent the possible game moves that can
be made by either player 1 or player 2. The
tree levels are built in a way that each player plays in turns and the
first level represents the moves made by player 1.  An
evaluation function \texttt{f(State)} is used to compute the score of
the board for each leaf of the tree. A node is a leaf when it can no longer be
expanded since the state represents a possible end of the game.  Finally, the
algorithm recursively minimizes or maximizes the scores of each node.
To select the best move for player 1, the
algorithm picks the move maximized at the root node.

In LM, the program starts with a root node (with the initial game state)
that is expanded with the available moves at each level. The graph of the
program is dynamic since nodes are created and eventually deleted once they
have nore more facts or no more references to them. The latter happens when the
leaves scores are computed or when a node fully minimizes or maximizes the
children scores. When the program ends, only the root node is have facts in its
database.

The LM code in Fig.~\ref{minimax:check-end} implements
the initial expansion of the tree. The first three rules (lines 1-10) deal
with the cases where no children nodes are created and the last three rules
(12-29) deal with cases where new nodes may be created.
Thes rule in lines 12-25 generate new nodes using the
\texttt{exists} language construct, which creates a new node address
\texttt{B}, where a new \texttt{play} fact is created with the new
board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
communicate with its parent.

Since our implementation
uses a FIFO ordering for nodes, meaning that when children are expanded, we
first expand each children and then the children of the children. The default
ordering of the system works well enough for most practical purposes. We note
however, that this is not optimal for this program since the complete tree needs
to be expanded before computing the scores at the leaves. This introduces memory
issues since, in the worst case, the space complexity of the program are
$\mathcal{O}(n)$, where $n$ is the number of nodes in the tree.

@@

expand(A, Board, [], N, P, \underline{Depth}),
N > 0
  -o maximize(A, N, -00, 0).

expand(A, Board, [], N, P, \underline{Depth}),
N > 0, NextPlayer <> RootPlayer
  -o minimize(A, N, +00, 0).

@@

With coordination, we set the priority nodes to be the same as the
depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads. In
Fig.~\ref{fig:minimax}, we have two threads (squares) and the first one has
expanded the tree on the left side. When a second thread comes up, it steals the
first unexpanded node and expands the tree from that position. Since threads
prioritize deeper nodes, the scores of the first leaves are immediatelly
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

@@

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{ \begin{tabular}[b]{ | c | c | c |}
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 11.80GB & 0.50MB \\ \hline
         2 & 12.19GB & 1.45MB \\ \hline
         4 & 13.82GB & 2.35MB \\ \hline
         8 & 14.87GB & 4.36MB \\ \hline
         16 & 13.79GB & 8.1MB \\ \hline
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=5cm]{results/min-max-tictactoe.png}}
   \end{center}
   \caption{Memory usage and scalability of the regular and coordinated versions
      of MiniMax.}
   \label{results:memory_minmax}
\end{figure}

>>>>

The algorithm proceeds by building a game tree, where each tree node represents a
game state and the children represent the possible game moves that can
be made by either player 1 or player 2. The
tree levels are built in a way that each player plays in turns and the
first level represents the moves made by player 1.  An
evaluation function \texttt{f(State)} is used to compute the score of
the board for each leaf of the tree. A node is a leaf when the game state can no longer be
expanded. Finally, the
algorithm recursively minimizes or maximizes the scores of each node.
To select the best move for player 1, the
algorithm picks the move maximized at the root node.

In LM, the program starts with a root node (with the initial game state)
that is expanded with the available moves at each level. The graph of the
program is dynamic since nodes are created and then deleted once they are no
longer needed. The latter happens when the
leaf scores are computed or when a node fully minimizes or maximizes the
children scores. When the program ends, only the root node has facts in its
database.

The LM code in Fig.~\ref{minimax:check-end} implements
the initial expansion of the tree. The first three rules (lines 1-10) deal
with the cases where no children nodes are created and the last three rules
(12-29) deal with cases where new nodes may be created.
The rule in lines 12-25 generate new nodes using the
\texttt{exists} language construct, which creates a new node address
\texttt{B}, where a new \texttt{play} fact is created with the new
board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
communicate with its parent.

As noted before in Section~\ref{sec:fifo}, we use a FIFO ordering for nodes,
meaning that when children are expanded, we first expand the children and
then the children of the children. We note
however, that this is not optimal for this program since the complete tree needs
to be expanded before computing the scores at the leaves. The memory
complexity of the program is $\mathcal{O}(n)$, where $n$ is the number of nodes
in the tree.

@@

expand(A, Board, [], N, P, \underline{Depth}),
N > 0, P = player1
  -o maximize(A, N, -00, 0).

expand(A, Board, [], N, P, \underline{Depth}),
N > 0, P = player2
  -o minimize(A, N, +00, 0).

@@

With coordination, we set the priority nodes to be the same as the
depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads. In
Fig.~\ref{fig:minimax}, we have two threads (squares) and the first one has
expanded the tree on the left side. When a second thread is idle, it steals the
first unexpanded node and expands the tree from that position. Since threads
prioritize deeper nodes, the scores of the first leaves are immediatelly
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

@@

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{\small\begin{tabular}[b]{ | c | c | c |}
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 11.80GB & 0.50MB \\ \hline
         2 & 12.19GB & 1.45MB \\ \hline
         4 & 13.82GB & 2.35MB \\ \hline
         8 & 14.87GB & 4.36MB \\ \hline
         16 & 13.79GB & 8.1MB \\ \hline
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=4.5cm]{results/min-max-tictactoe.png}}
   \end{center}
   \caption{Memory usage and scalability of the regular and coordinated versions
      of MiniMax.}
   \label{results:memory_minmax}
\end{figure}

In Fig.~\ref{results:memory_minmax} we compare the memory usage and scalability
of the coordinated MiniMax against the regular MiniMax. The coordinated version
sees a huge reduction of memory usage, needing at most 8MB for 16 threads, while
the regular version needs around 14GB. Note that as the number of threads goes
up, memory usage also goes up. This is an artifact of our parallel memory
allocator that allocates large chunks of memory before hand.
In terms of run time, our experimental results show linear scalability for the coordinated
version and a 20\% run time reduction when using 16 threads.

### 2014-11-05 splash-bp.tex
More corrections.

Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
where nodes exchange messages with their immediate neighbors and apply some computations to the messages
received.

@@

// SECOND PHASE
second-phase(A, [], _)
   -o *underline{set-priority(A, 0.0)}, waiting(A), update(A).
second-phase(A, [A], Back)
   -o update(A), waiting(Back),
      waiting(A), *underline{set-priority(A, 0.0)}.
second-phase(A, [A, B | Next], Back)
   -o update(A), waiting(Back), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
\end{Verbatim}
  \caption{Coordination code for the Splash Belief Propagation program. When a
     linear fact is prefixed by \texttt{@}, it indicates that the fact is going
  to be re-derived.}
  \label{code:sbp}
\end{figure}
\normalsize

>>>>

Randomized and approximation algorithms can obtain significant benefits from
coordination directives because although the final program results will not be
exact, they follow important statistical properties and can be computed faster.
Examples of such programs are Loopy Belief Propagation.
Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
where nodes exchange messages with their immediate neighbors and apply some computations to the messages
received.

@@

// SECOND PHASE
second-phase(A, [], _)
   -o *underline{set-priority(A, 0.0)}, waiting(A), update(A).
second-phase(A, [A], Back)
   -o update(A), waiting(Back),
      waiting(A), *underline{set-priority(A, 0.0)}.
second-phase(A, [A, B | Next], Back)
   -o update(A), waiting(Back), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
\end{Verbatim}
  \caption{Coordination code for the Splash Belief Propagation program. LM needs
     50 lines of rules to implement splash trees, while GraphLab needs a
     total of 350 lines of C++ to implement the same functionality.  Note
     that when a linear fact is prefixed by \texttt{@}, it indicates that the
     fact is going to be re-derived.}
  \label{code:sbp}
\end{figure}
\normalsize
